---
title: "DESeq_test_scTE"
output: html_document
date: '2024-10-22'
---
# this notebook serves as a template to run DESeq and make volcano plots for differential genes
- followed tutorial from [Deseq2 manual](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## import 
```{r}


if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.19")

BiocManager::install("DESeq2")

BiocManager::install('EnhancedVolcano')






library( DESeq2)
library(EnhancedVolcano)

```

## load data
- the count matrix should be a gene x sample matrix 
- the col data should be a table that as the sample names as indices ( samples matching the count matrix) and experimental conditions in each column
- should be un-normalized integer values ( raw count table)

```{r}
getwd()
project_dir = '~/Dropbox/scTE_try/deseq_TE_files/'

cts = read.csv( paste0( project_dir , 'deseq_input_cts.csv' ), header = TRUE,row.names = 1, check.names = FALSE )
coldata =  read.csv( paste0( project_dir , 'deseq_input_coldata.csv' ), stringsAsFactors = TRUE) 

cts2 = read.csv( paste0( project_dir , 'deseq_input_raw_cts.csv' ), header = TRUE,row.names = 1, check.names = FALSE )
coldata2 =  read.csv( paste0( project_dir , 'deseq_input_raw_coldata.csv' ), stringsAsFactors = TRUE) 

#
```

## run DESeq 
```{r}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design= ~Patient_ID + Sample_Classification)

dds <- DESeq(dds)

resultsNames(dds) # lists the coefficients
res_all = results(dds)

# get specific results: contrast should have 3 arguments: the colname of the condition compared, and the two labels for the condition

res1  <- results(dds, contrast = c("Sample_Classification", "AD", "NL"))
res2  <- results(dds, contrast = c("Sample_Classification", "SER", "NL"))
res3  <- results(dds, contrast = c("Sample_Classification", "SER", "AD"))
# or to shrink log fold changes association with condition:
#res <- lfcShrink(dds, coef="Sample_Classification_NL_vs_AD", type="apeglm")

filtered_result_plot_path = '~/Dropbox/scTE_try/deseq_TE_files/filtered_res_plots/'

```

## quick visualize result

check overall result
```{r}
head( res_all) # this is for SER VS AD
summary( res_all)
```
compare if res3 and res_all are the same
- look like to be the same just different direction
```{r}
head(res3)
```
```{r}
head( res_all)
```

## Make vocano plot for the comparison

Here's the function
Updated 5/9/25
```{r}
customize_vocano = function(plot_res, 
                            positive_color = 'salmon', negative_color = 'royalblue', ns_color = 'gray',
                            pval_colname = 'pvalue', lfc_colname = 'log2FoldChange', 
                            pval_lim = 0.05, positive_LFC_lim = 0.5, negative_LFC_lim = -0.5, 
                            contrast_name = c('group1' ,'group2'), 
                            labSize = 5, drawConnectors = TRUE  ){
  # this function makes a vocano plot from DESeq2 result, customizing the color scheme and thresholding limits for the LFC and p-vals
  # Args:
  #   plot_res is the output from DESeq2 ( plot_res = result(dds))
  #   positive_color, negative_color, ns_color are the colors used for the significant genes that are up, down or ns regulated in the comparison group
  #   pval_colname and lfc_colname are the column names from the dds result that would be used for the plot. Defaults are pvale and log2FolkdChange, but some times, might prefer padj as pvalues
  #   pval_lim, positive_LFC_lim, negative_LFC_lim are numerical values for the thresholds 
  #   positive_LFC_lim and  negative_LFC_lim are the log2 fold change thresholds 
  #   contrast_name should be a vector of two strings for the compared variable names in the same order as DESeq output ( eg. ['SER', 'AD'])
  # return:
  #   the vocano plot
  
  #creating a color list for each data points
  print( "debug start")
  keyvals = ifelse( 
  (plot_res[[lfc_colname]] < negative_LFC_lim & plot_res[[pval_colname]] < pval_lim), negative_color, 
  ifelse( (plot_res[[lfc_colname]]> positive_LFC_lim & plot_res[[pval_colname]] < pval_lim), positive_color,
          ns_color))
  
  print( "got pos and neg keywords")
  keyvals[is.na(keyvals)] = ns_color
  
  print( "got NS keywords")
  names(keyvals)[keyvals == positive_color] <- paste( contrast_name[1], 'High')
  names(keyvals)[keyvals == ns_color] <- 'NS'
  names(keyvals)[keyvals == negative_color] <- paste( contrast_name[2], 'High')
  
  print( "come through coloring")
  
  # make the plot
  p = EnhancedVolcano(plot_res,
    lab = rownames(plot_res),
    x = lfc_colname,
    y = pval_colname,
    title = paste( contrast_name[1], 'vs', contrast_name[2] ), 
    
    FCcutoff = positive_LFC_lim, 
    pCutoff = pval_lim, 
    pointSize = 1.5,
    
    colCustom = keyvals,
    colAlpha = 0.7,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 3,
    labSize = labSize,
    
    drawConnectors = drawConnectors,
    widthConnectors = 0.2,
    max.overlaps = 15
    )
  
  return(p)
  
}

```


draft code to make the plot function
```{r}

keyvals = ifelse( 
  (plot_res$log2FoldChange < -0.5 & plot_res$pvalue <0.05), 'royalblue', 
  ifelse( (res$log2FoldChange > 0.5 & plot_res$pvalue <0.05), 'salmon',
          'gray'))
keyvals[is.na(keyvals)] = 'gray'
names(keyvals)[keyvals == 'salmon'] <- 'SER high'
names(keyvals)[keyvals == 'gray'] <- 'NS'
names(keyvals)[keyvals == 'royalblue'] <- 'AD High'


EnhancedVolcano(res_all,
    lab = rownames(res_all),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'SER vs AD', 
    
    FCcutoff = 0.5, 
    pCutoff = 0.05, 
    pointSize = 1,
    
    colCustom = keyvals,
    colAlpha = 0.6,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 3
    )
```

## test the function
```{r}
p = customize_vocano( res_all,  contrast_name = c('SER', 'AD'), labSize = 8)

p

#ggsave( "~/Dropbox/scTE_try/Deseq_SER_AD.png", plot = p, width = 10, height =8, dpi = 300 )
#ggsave( "~/Dropbox/scTE_try/Deseq_SER_AD_large_label.png", plot = p, width = 10, height =8, dpi = 300 )

```



## results

AD vs NL
```{r}
summary(res1)
head( res1) #AD vs NL 
```

```{r}

# a no label version
p2 = customize_vocano(res1, pval_lim = 0.05, positive_LFC_lim = 2, negative_LFC_lim = -2,   contrast_name = c( 'AD', 'NL'), positive_color = 'royalblue', negative_color = 'orange', labSize = 0, drawConnectors = FALSE )
p2

#ggsave( "~/Dropbox/scTE_try/Deseq_AD_NL.png", plot = p2, width = 10, height =8, dpi = 300 )
ggsave( paste0( filtered_result_plot_path, "Deseq_AD_NL.png" ), plot = p2, width = 10, height =8, dpi = 300 )


# a label version
p2 = customize_vocano(res1, pval_lim = 0.05, positive_LFC_lim = 2, negative_LFC_lim = -2,   contrast_name = c( 'AD', 'NL'), positive_color = 'royalblue', negative_color = 'orange', labSize = 5, drawConnectors = TRUE )
p2

#ggsave( "~/Dropbox/scTE_try/Deseq_AD_NL.png", plot = p2, width = 10, height =8, dpi = 300 )
ggsave( paste0( filtered_result_plot_path, "Deseq_AD_NL_label.png" ), plot = p2, width = 10, height =8, dpi = 300 )

```



## save DGE result
```{r}
# SER vs AD
write.csv(as.data.frame(res3),  file="Deseq_SER_AD_results.csv")

# SER vs NL
#res2
write.csv(as.data.frame(res2),  file="Deseq_SER_NL_results.csv")


# AD vs NL
#res1
write.csv(as.data.frame(res1),  file="Deseq_AD_NL_results.csv")

getwd()
```

